Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-19 paper.tex
First pass at the discussion

\section{Discussion}
The COVID-19 pandemic continues to be a global emergency,
with persistent high levels of infection. This high prevalence has allowed the
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
higher risk of coinfection, increasing opportunities for new phenotypically
distinct recombinants to emerge and spread.

In this proof-of-concept study, we demonstrate the feasibility of
reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
that is simply not possible using other available ARG inference methods
\citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}. We have accomplished
this feat by building on top of recent advances in storing and processing large
ARGs (tskit) and inferring genealogical relationships efficiently at scale
(tsinfer). The highly compact representation of the ARGs as succinct tree
sequences, coupled with vectorised computing technologies, enables the large
SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and
allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of
data. We believe that further development of the sc2ts methodology and software
can create a powerful tool (1) to efficiently detect emerging or previously
overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and
recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3)
to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with
unprecedented global sequencing coverage.

ARGs are rich genealogical structures that contain information about the
genetic relationships among sampled genomes in the presence of recombination.
Here, we have conducted several analyses of the preliminary ARGs to show that
the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
among major SARS-CoV-2 lineages; (2) recover biologically plausible
recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
SARS-CoV-2 evolutionary history.

ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
package. However, we have noted several areas of differences (e.g., poorly
calibrated branch lengths), which are the focus of future development of sc2ts.
We believe that some of these differences can be reduced by improving and
refining our approach and strategy to handle samples with erroneously reported
collection dates (i.e., time travellers that have passed the maximum submission
delay of 30 days) and to modify the topology of daily trees.

Our results show that sc2ts identifies the early recombinants (pre-2022)
previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
Pango recombinants, which are designated after community perusal of supporting
genomic evidence. We have found several previously unreported and overlooked
recombinant sequences in the wide ARG. After further improvements and
enhancements, we will use sc2ts to build a comprehensive ARG to search for
recombination signals among all the 15 million and counting SARS-CoV-2 genomes
available.

``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.'' Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.

\section{Improvements}
Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.

\begin{itemize}
\item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs. Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out ``time travellers'', but the sampling time can still be wrong by days or tens of days. ***describe possible approaches to correct this?***
\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.). An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
\item We have used a single mismatch value in all the HMM runs, set so that having three mutations is as likely as having a single recombination. This parameter choice is arbitrary, and higher values may reduce the number of origins for those Pango lineages that were inferred to descend from more than one recombinant ancestor. In addition, because both the mutation rate and recombination rate varies across the SARS-CoV-2 genome, it may be helpful to allow the mismatch value to vary along the genome.
\item By treating mutations equally in the current HMM implementation, we implicitly assume that the rates of different types of mutations are identical. However, as reported previously \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide substitution models (e.g., allowing for different rates of transition and transversion) can be incorporated into HMM likelihood calculations.
\item Our current masking strategy handles indels as missing data. Doing this can remove information helpful to detect recombinants. For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
\item Daily trees are built using the UPGMA algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm to build daily trees would leave these "soft polytomies" in, removing the need for the un-resolving step.
\end{itemize}

\section{Conclusions}
The global understanding of viral evolution during the COVID-19 pandemic is
unprecedented, with over 15 million sequences shared to date. We envisage that
once fully developed and tested, sc2ts methodology and software may provide a
unified computational framework to study the evolution and epidemiology of
SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
exemplify the numerous applications possible using the ancestral recombination
graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
and virologists to join us and the broader tskit community to build the most
complete representation of the evolutionary history of SARS-CoV-2. Moreover,
while we have focussed on the particular application to SARS-CoV-2, the methods
we describe are readily applicable to the analysis of many other viral
pathogens. Solving the challenges in building and analysing large viral
genealogies would therefore contribute to the important goal of improving
preparedness for potential future pandemics.

>>>>

\section{Discussion}
% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
% unprecedented opportunity to study the evolution of a virus. Even if the
% public health emergency is officially over, it doesn't mean it's not coming
% back and we should be prepared. Recombination is now a vital factor, and
% present a way in which it can be incorporated. Our results are accurate.

The COVID-19 pandemic continues to be a global emergency [DOES IT? Didn't the
WHO downgrade recently?]
with persistent high levels of infection. This high prevalence has allowed the
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
higher risk of coinfection, increasing opportunities for new phenotypically
distinct recombinants to emerge and spread.
[Handling recombination properly as an ongoing process is vital. Our results
here show that recombination can be incorporated into a cohesive joint
model of viral evolution. We basically recapitulate what is known, plus
there's a whole bunch of other stuff that we haven't explored because
we're in the validation stage.]

% PARA 2. Here are things that could be improved in the short term. Here's a
% few more, longer term ideas.
Nonetheless, \texttt{sc2ts} is currently ``alpha'' quality software, and
there are some relatively straightforward changes that could be made to
improve inference quality. Firstly, the pipeline used to identify and
mask erroneous sites in the input alignments is simplistic, and, among
other issues, results in multi-base indels being marked as missing data
(Section~\ref{sec:data_preprocessing}). A more sophisticated
approach~\citep[e.g.][]{Aksamentov2021-hj} would
likely yield significant improvements, and reduce the effect sites with
artefactually high levels of mutation (e.g., site 28271;
Section~\ref{sec:breakpoint_intervals}).
Using a pre-existing tree built using state-of-the-art phylogenetics methods
for the early stages of the pandemic (Section~\ref{sec:backbone_phylogeny}),
and minor adaptations to standard node-dating methods
(Section~\ref{sec:node_dating}) should help resolve the most notable issues with
the inferred backbone phylogeny (Figure~\ref{fig:cophylogeny}).
Trees constructed from daily sample clusters have a surprisingly large
influence on the overall ARG topology
(Section~\ref{sec:sample-cluster-tree-inference}), and so using a more
sophisticated tree building approach should yield clear improvements.
The unrealistically large number of reversion mutations
(Section~\ref{sec:mutation_spectrum}) may be reduced by
improvements to the current parsimony
hueristics (Section~\ref{sec:parsimony-heuristics}).
A major source of errors are the ``time-traveller'' samples, whose
provided collection dates are months (or years) too early
(Section~\ref{sec:filtering_time_travellers}).
% Hmm, not sure about this. Wouldn't this also rule out new major lineages
% also (which tend to show up with a bunch of new mutations?) I guess it
% depends on the threshold, and we could set it according to previous
% experience with new lineages?
While it is unclear how
we might solve this problem in general, some simple expedients such
as filtering out sequences that exceed a given cost in the LS HMM
(i.e., number of mutations and recombination switches) may work well
in practise. Such an approach would also reduce the impact of sequences
with high levels of sequencing error (which contribute a large number of
mutations, currently).
Taken together, these and other relatively minor improvements should
enable inference over much larger subsets of the dataset, and
give a clearer picture of the combined processes of recombination and
mutation over the pandemic so far.

% What are the more blue-sky type things? Basically these are things we don't
% really want to do for the first "real" version of sc2ts.
An attractive feature of \texttt{sc2ts} is that the most difficult part of
the inference problem---finding likely recombinant paths through the existing ARG
for new samples---is solved exactly under a well-defined statistical model,
using established HMM methodology (Section~\ref{sec:ls}).
The implementation currently uses
a single, arbitrarily chosen, maximum likelihood path via the
Viterbi algorithm, but there are numerous ways in which the HMM machinery
could be extended in order to explore the set of possible paths, or to
quantify the uncertainty around it. Similarly, the current parameterisation
of the HMM with a single mismatch ratio is simplistic, and it would
be straightforward to condition on per-site mutation rates. (Per nucleotide
state transition, is also possible, with some additional development.)
Recombination breakpoints for the ARG are currently inserted at the
right-most extent of the possible interval
(Section~\ref{sec:breakpoint_intervals}). More likely locations for the
breakpoint could be chosen within the interval, for example based
on the sequence motifs.

% PARA 3. Handling the data is hard, so methods for making this easier
% are important. Classical phylo approaches don't scale to millions.
% The ability to quickly develop new analyses, at scale, will be crucial.
% Sc2ts is largely composed of other parts being reused, with mainly being
% concerned with masking input alignments. We wrote no C at all. This kind
% of code reuse and ease of analysis is a major bonus.
The vast volumes of whole genome sequence data generated during the pandemic
has presented classical phylogenetics methods and software with major difficulties.
Standard interchange formats such as FASTA, Newick and VCF were simply not designed
to deal with millions of samples, and their limitations have come sharply
into focus~\citep{Turakhia2021-ur,de2023maximum}.
[Finish this up with a discussion of how easy it is to do stuff with our
ARGS, how awesome tskit is, and how little code there is in sc2ts.
Being able to quickly develop new methods by putting together high
performance, high quality components is an essential element of
preparedness for futture pandemics.]

% The COVID-19 pandemic continues to be a global emergency,
% with persistent high levels of infection. This high prevalence has allowed the
% proliferation of many variants, with more than 600 Pango-designated lineages
% circulating globally in the last three months (January to March, 2023;
% https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
% higher risk of coinfection, increasing opportunities for new phenotypically
% distinct recombinants to emerge and spread.

% In this proof-of-concept study, we demonstrate the feasibility of
% reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
% that is simply not possible using other available ARG inference methods
% \citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}. We have accomplished
% this feat by building on top of recent advances in storing and processing large
% ARGs (tskit) and inferring genealogical relationships efficiently at scale
% (tsinfer). The highly compact representation of the ARGs as succinct tree
% sequences, coupled with vectorised computing technologies, enables the large
% SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and
% allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of
% data. We believe that further development of the sc2ts methodology and software
% can create a powerful tool (1) to efficiently detect emerging or previously
% overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and
% recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3)
% to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with
% unprecedented global sequencing coverage.

% ARGs are rich genealogical structures that contain information about the
% genetic relationships among sampled genomes in the presence of recombination.
% Here, we have conducted several analyses of the preliminary ARGs to show that
% the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
% among major SARS-CoV-2 lineages; (2) recover biologically plausible
% recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
% SARS-CoV-2 evolutionary history.

% ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
% SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
% sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
% a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
% package. However, we have noted several areas of differences (e.g., poorly
% calibrated branch lengths), which are the focus of future development of sc2ts.
% We believe that some of these differences can be reduced by improving and
% refining our approach and strategy to handle samples with erroneously reported
% collection dates (i.e., time travellers that have passed the maximum submission
% delay of 30 days) and to modify the topology of daily trees.

% Our results show that sc2ts identifies the early recombinants (pre-2022)
% previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
% Pango recombinants, which are designated after community perusal of supporting
% genomic evidence. We have found several previously unreported and overlooked
% recombinant sequences in the wide ARG. After further improvements and
% enhancements, we will use sc2ts to build a comprehensive ARG to search for
% recombination signals among all the 15 million and counting SARS-CoV-2 genomes
% available.

% ``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.'' Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.

% \section{Improvements}
% Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.

% \begin{itemize} \item A potential source of error in sc2ts ARGs comes from
% incorrectly reported collection dates of the samples, which can cause sample
% genomes to be attached to the wrong parts of the ARGs. Thus far, we have
% attempted to correct the most offending errors in the ARGs by filtering out
% ``time travellers'', but the sampling time can still be wrong by days or tens
% of days. ***describe possible approaches to correct this?*** \item To avoid
% time travellers, we used a maximum submission delay, but this will
% systematically bias against submissions from countries with greater submission
% delays (Kalia et al.). An area of future work is to add back the excluded
% sequences to a built ARG to assess which ones are likely true time travellers.
% \item We have used a single mismatch value in all the HMM runs, set so that
% having three mutations is as likely as having a single recombination. This
% parameter choice is arbitrary, and higher values may reduce the number of
% origins for those Pango lineages that were inferred to descend from more than
% one recombinant ancestor. In addition, because both the mutation rate and
% recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
% allow the mismatch value to vary along the genome. \item By treating mutations
% equally in the current HMM implementation, we implicitly assume that the rates
% of different types of mutations are identical. However, as reported previously
% \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in
% SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide
% substitution models (e.g., allowing for different rates of transition and
% transversion) can be incorporated into HMM likelihood calculations. \item Our
% current masking strategy handles indels as missing data. Doing this can remove
% information helpful to detect recombinants. For example, sequences of the Pango
% recombinant XP were not identified as recombinants, because XP’s characteristic
% multibase deletion was masked. \item Daily trees are built using the UPGMA
% algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of
% the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm
% to build daily trees would leave these "soft polytomies" in, removing the need
% for the un-resolving step. \end{itemize}

% \section{Conclusions}
% The global understanding of viral evolution during the COVID-19 pandemic is
% unprecedented, with over 15 million sequences shared to date. We envisage that
% once fully developed and tested, sc2ts methodology and software may provide a
% unified computational framework to study the evolution and epidemiology of
% SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
% exemplify the numerous applications possible using the ancestral recombination
% graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
% and virologists to join us and the broader tskit community to build the most
% complete representation of the evolutionary history of SARS-CoV-2. Moreover,
% while we have focussed on the particular application to SARS-CoV-2, the methods
% we describe are readily applicable to the analysis of many other viral
% pathogens. Solving the challenges in building and analysing large viral
% genealogies would therefore contribute to the important goal of improving
% preparedness for potential future pandemics.
